---
title: "Analysis of Sustainable Development Goals"
author: "SOfia"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction
This analysis focuses on the first three Sustainable Development Goals (SDGs) SDG1(Ending Poverty),SDG2(Zero Hunger),SDG3(Good Health and Well-being in Kenya,Ghana,EGypt,Botswana and Cameroon.

## Data Import
```{r data-import}
# Load necessary libraries
library(tidyverse)
library(table1)

# Load the dataset
gdp_data <- read.csv("Final.csv")

# Preview the dataset
head(gdp_data)
#View(gdp_data)
```

## Data Cleaning
I extracted the data from a set of datasets spanning the years 2000 to 2017 for the specified countries. Due to the presence of many missing values, i replaced them with the mean value of the corresponding variables for each country. In total, there were 90 observations and 10 variables.
```{r data-cleaning}

# Display cleaned dataset summary
summary(gdp_data)
str(gdp_data)
```
##factor of categorical variables  
```{r}

gdp_data$Code<-factor(gdp_data$Code)
gdp_data$year<-factor(gdp_data$year)
str(gdp_data)
View(gdp_data)

```

  
```{r}
#data$Year<-factor(data$Year)
gdp_data$GDPphead<-as.numeric(gdp_data$GDPphead)
gdp_data$NeonatalDeathRate<-as.numeric(gdp_data$NeonatalDeathRate)
gdp_data$NoPeopleBPovertyLine<-as.integer(gdp_data$NoPeopleBPovertyLine)
gdp_data$unfedprcnt<-as.numeric(gdp_data$unfedprcnt)
gdp_data$stuntPerc5<-as.numeric(gdp_data$stuntPerc5)
gdp_data$stuntPerc5 <- round(gdp_data$stuntPerc5, 2)
gdp_data$GDPphead <- round(gdp_data$GDPphead, 2)
gdp_data$NeonatalDeathRate <- round(gdp_data$NeonatalDeathRate, 2)
gdp_data$unfedprcnt <- round(gdp_data$unfedprcnt, 2)
```

 
```{r}
#clean_data <- na.omit(gdp_data)
# Preview cleaned data
str(gdp_data)
```

  


## Descriptive Statistics
Summarize maternal mortality and GDP per capita across regions and countries.

```{r descriptive-stats}
# Summary statistics for maternal mortality and GDP
attach(gdp_data)
tab2<- table1(~ GDPphead+MaternalMortRate|Code , data=gdp_data)
tab2
```
The table shows stunting is highest in Cameroon (34.4\%) and poverty in Kenya (62.0\%), while Egypt has the lowest rates for both. Botswana has the highested percentage (21.8\%), and neonatal deaths range from 16.6 in Egypt to 32.0 in Cameroon. Overall, health and socioeconomic indicators vary significantly across countries.


## Exploratory Data Analysis
Visualize the relationship between maternal mortality and GDP per capita, and examine distributions.  
```{r}

 par(mfrow=c(2,2))
 hist(GDPphead,col="lightgreen",breaks = 20)
 hist(MaternalMortRate,col="pink",breaks=20)
 boxplot(GDPphead~Code,col="lightgreen")
 boxplot(GDPphead~year)
```


## Modeling: Maternal Mortality and GDP per Capita
```{r modeling-maternal-mortality}
# Fit the linear model
model1 <- glm(MaternalMortRate ~ GDPphead, data = gdp_data,family = "poisson")

# Display model summary
summary(model1)


```
## plots of the model  
```{r}

plot(model1)
```

## Exploratory data Analysis  
```{r}
 par(mfrow=c(2,2))
 hist(stuntPerc5,col="lightgreen",breaks = 20)
 hist(PrcntBelowPovertyLine,col="pink",breaks=20)
 hist(NeonatalDeathRate,col="pink",breaks=20)
 hist(unfedprcnt,col="pink",breaks=20)
 
 par(mfrow=c(2,2))
 boxplot(unfedprcnt~Code,col="lightgreen")
 boxplot(unfedprcnt~year)
 boxplot(PrcntBelowPovertyLine~Code,col="lightgreen")
 boxplot(PrcntBelowPovertyLine~year)
 par(mfrow=c(2,2))
 boxplot(stuntPerc5~Code,col="lightgreen")
 boxplot(stuntPerc5~year)
 
```

## Summary Table  
```{r}
tab3<- table1(~ stuntPerc5+PrcntBelowPovertyLine+unfedprcnt+NeonatalDeathRate|Code , data=gdp_data)
tab3
```



## Second model  
```{r}
model2<-lm(NeonatalDeathRate~stuntPerc5+PrcntBelowPovertyLine+unfedprcnt,data = gdp_data)
summary(model2)
```
## model evaluation  
```{r}
plot(model2)
```


---

